---
title: "Drug Related Deaths in Scotland 2017"
author: "Christian Cole"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(readxl)
library(reshape2)
library(ggplot2)
library(tibble)
library(dplyr)

# default region colours
reg.colours = c("#4D4D4D", "#AEAEAE")

# The colour-blind palette with grey:
myPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

# read data - download first as read_excel() can't do URLs
xlfile = tempfile(fileext = '.xlsx')
download.file("https://www.nrscotland.gov.uk/files//statistics/drug-related-deaths/17/drug-related-deaths-17-tabs-figs.xlsx", xlfile, mode="wb")

```

## Tab {.tabset}

### Table 2

```{r t2}
# read data
table8.dat = read_excel(xlfile, sheet = 4, col_names=FALSE)

# extract core data and convert to numbers
dat = table8.dat[36:43,c(1,3:7)]
dat = data.frame(apply(dat, 2, as.numeric))

# add colnames
colnames(dat) <- c('Year','Drug Abuse','Accidental Poisoning','Suicide','Assault by Drugs','Undetermined')

# remove averages row and save
mean13.17 = dat[8,]
dat = dat[-8,]

# factorise the years to turn into a label and melt
dat$Year <- factor(dat$Year)
dat.m = melt(dat)

year.totals = dat.m %>% group_by(Year) %>% summarise(Total = sum(value))

ggplot(dat.m, aes(x=Year, y=value, fill=variable)) + 
  ggtitle("Underlying Cause of Death") +
  ylab("No. Deaths") +
  geom_col() +
  geom_text(aes(x=Year, y=Total+28, label = Total, fill = NULL), data=year.totals, size=2.8) +
  scale_fill_manual(values = myPalette) +
  theme_minimal() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.title = element_blank())

```

### Table 3

```{r t3, fig.width=7}
# read data
table.dat = read_excel(xlfile, sheet = 5, col_names=FALSE)

# extract core data and convert to numbers
dat = table.dat[15:32,c(1,3:14)]
dat = data.frame(apply(dat, 2, as.numeric))

colnames(dat) <- c('Year','Heroin/morphine','Methodone','Heroin or OST','Codeine', 'Dihydrocodeine', 'Any opiate', 'Any benzodiazepine', 'Diazepam', 'Cocaine', 'Ecstasy', 'Amphetamines', 'Alcohol')

# combine the codeine/dihydrocodeine cols and remove
dat = data.frame(dat, Codeines = dat$Codeine + dat$Dihydrocodeine)

# subtract the diazepam data from 
# the any benzo data to make an "other benzo" column
dat = data.frame(dat, "Other benzodiazepines"=dat$Any.benzodiazepine - dat$Diazepam)

# remove uninterest columns:
#   heroin or OST and any opiate are not very informative
#   Codeine/dihydrocodeine unnecessary now as they've been combined
#   any benzo is superceded by other benzos
# individual cases)
dat = dat[,c(1:3,9:15)]

opiates = dat[,c(1:4,9:10)]
others = dat[,c(1,5:8)]

# factorise the years to turn into a label and melt
opiates$Year <- factor(opiates$Year)
opiates.m = melt(opiates)

ggplot(opiates.m, aes(x=Year, y=value, group=variable, colour=variable)) +
  ggtitle("Drugs Implicated in Death", subtitle = 'Opiates and benzodiazepines') +
  ylab("No. of Deaths") +
  ylim(c(0,500)) +
  geom_vline(xintercept=which(opiates$Year == '2008'), colour="grey", linetype="dashed") +
  geom_line(size=1.1) +
  scale_color_manual(values = myPalette, labels = c('Heroine or Morphine','Methadone','Diazepam','Codeines','Other\nbenzodiazepines')) +
  scale_x_discrete(breaks = seq(2000,2016,2)) +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        legend.title = element_blank())

# factorise the years to turn into a label and melt
others$Year <- factor(others$Year)
others.m = melt(others)

ggplot(others.m, aes(x=Year, y=value, group=variable, colour=variable)) +
  ggtitle("Drugs Implicated in Death", subtitle = 'Others') +
  ylab("No. of Deaths") +
  ylim(c(0,500)) +
  geom_vline(xintercept=which(opiates$Year == '2008'), colour="grey", linetype="dashed") +
  geom_line(size=1.1) +
  scale_color_manual(values = myPalette) +
  scale_x_discrete(breaks = seq(2000,2016,2)) +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        legend.title = element_blank())


```

### Table 4

Table 4 describes quite a lot of data represented in different ways. Each is presented as a different plot.

```{r t4a}
# read data
table.dat = read_excel(xlfile, sheet = 6, col_names=FALSE)

# extract core data and convert to numbers  there are some 
# columns used as fillers. Gah!
dat = table.dat[14:31,c(1,2,4,5,7:13,15:17)]
dat = data.frame(apply(dat, 2, as.numeric))

colnames(dat) <- c('Year','all','male','female','u14','15_24','25_34','35_44','45_54','55_64','o65','lq','median','uq')

ggplot(dat, aes(x=Year, y=median)) + 
  ggtitle("Median Age at Death", subtitle = 'Shaded area represents the lower to upper quartile range of ages.') +
  ylab("Age") +
  geom_line(size=1.2) +
  ylim(c(0,50)) +
  stat_summary(ymin=dat$lq, ymax=dat$uq, geom="ribbon", alpha=0.2, fill='red') +
  scale_x_continuous(breaks = seq(2000,2017,2)) +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        legend.title = element_blank())
```



Median age of drug-related death is increasing from 30 in 2000 to 41 in 2017.

```{r t4b}
# calc percentage m/f deaths
dat = data.frame(dat, fempc = 100*dat$female/dat$all, malpc = 100*dat$male/dat$all)

dat$Year = factor(dat$Year)
dat.sex = dat[,c('Year','fempc','malpc')]
dat.sex.m = melt(dat.sex)

ggplot(dat.sex.m, aes(x=Year, y=value, group=variable, colour=variable)) + 
  ggtitle("Percentage of Male vs Female Deaths") +
  ylab("Percentage Deaths") +
  geom_line(size=1.2) +
  ylim(c(10,90)) +
  scale_color_discrete(name='Sex', breaks=c('malpc','fempc'),labels=c('Male','Female')) +
  scale_x_discrete(breaks = seq(2000,2017,2)) +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        legend.title = element_blank())
```

Males are mostly affected by drug-related deaths, but females are increasing.

```{r t4c}
## get data from lower table

calc_percent <- function(x) {
  year = x[1]
  x = x[-1]
  total = sum(x)
  pc.x = round(100*x/total,2)
  return(c(year, pc.x))
}

# read data
table.dat = read_excel(xlfile, sheet = 6, col_names=FALSE)

# extract male data and convert to numbers  there are some 
# columns used as fillers. Gah!
mdat = table.dat[40:57,c(1,4,5,7:9)]
mdat = data.frame(apply(mdat, 2, as.numeric))

# turn into percentages
mdat = data.frame(t(apply(mdat,1,calc_percent)))
colnames(mdat) <- c('Year','u24','25-34','35-44','45-54','o55')

# extract female data and convert to numbers  there are some 
# columns used as fillers. Gah!
fdat = table.dat[40:57,c(1,12:13,15:17)]
fdat = data.frame(apply(fdat, 2, as.numeric))

# turn into percentages
fdat = data.frame(t(apply(fdat,1,calc_percent)))
colnames(fdat) <- c('Year','u24','25-34','35-44','45-54','o55')

dat = rbind(mdat,fdat)

dat = data.frame(dat, Sex=rep(c('Male','Female'),each=nrow(mdat)))

dat$Year = factor(dat$Year)
dat.m = melt(dat)

ggplot(dat.m, aes(x=Year, y=value, fill=variable)) +
  ggtitle('Deaths by Sex and Age') +
  ylab('Percentage Deaths') +
  geom_col() +
  facet_grid(Sex ~ .) +
  scale_fill_manual(name='Age', labels=c('under 25','25-34','35-44','45-54', 'over 55'), values = myPalette) +
  scale_x_discrete(breaks = seq(2000,2017,2)) +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        strip.background = element_rect(fill='grey'),
        legend.title = element_blank())


```

In 2000 deaths were largely in the under 35s. In 2017 they are dominated by the 35 and over. Particluarly in females.

### Table 5

```{r t5, fig.width=5, fig.height=6.5}
# read data
table.dat = read_excel(xlfile, sheet = 7, col_names=FALSE)

# extract core data and convert to numbers
dat = table.dat[c(22:26,29:33),c(1,3:5,7)]
#dat = data.frame(apply(dat, 2, as.numeric))
colnames(dat) <- c("Age",'Drug Abuse','Accident','Suicide','Undetermined')
dat = data.frame(dat, Sex = rep(c("Male","Female"), each = 5))
dat.m = melt(dat, id=c('Age','Sex'))
dat.m$value = as.numeric(dat.m$value)
dat.m$Age = factor(dat.m$Age, levels = c('Under 25','25-34','35-44','45-54','55 and over'))

age.totals = dat.m %>% group_by(Age,Sex) %>% summarise(Total = sum(value))


ggplot(dat.m, aes(x=Age, y=value, fill=variable)) + 
  ggtitle("Drug-Related Deaths in Scotland", subtitle="By Sex, Age and Underlying Cause of Death") +
  ylab('No. Deaths') +
  geom_col() +
  facet_grid(Sex ~ .) +
  scale_fill_manual(values = myPalette) +
  geom_text(aes(x=Age, y=Total+14, label = Total, fill = NULL), data=age.totals, size=2.8) +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        strip.background = element_rect(fill='grey'),
        legend.title = element_blank())



```


### Table 8

```{r t8}
# read data
table8.dat = read_excel(xlfile, sheet = 10, col_names=FALSE)

# extract core data and convert to numbers
dat = table8.dat[10:27,2:6]
dat = data.frame(apply(dat, 2, as.numeric))

# extract mean for last five years
yr5mean = as.numeric(table8.dat[29,7])

# add labels
names(dat) <- c("15-24","25-34","35-44","45-54","55-64")
dat = cbind(Year=factor(seq(2000,2017)), dat)

# melt data and plot
# note: dashed line is for average deaths 2012-2016 for ages 15-64
dat.m = melt(dat, variable.name = "Age")
ggplot(dat.m, aes(x=Year, y=value, group=Age, colour=Age)) + 
  geom_hline(yintercept = yr5mean, colour="grey", linetype="dashed") + 
  geom_line(size=1.2) + 
  ylab("Deaths per 1000 population") + 
  annotate("text", label="Mean 2013-2017", x=16.8,y=yr5mean-0.015,size=3,colour="grey") +
  scale_x_discrete(breaks=seq(2000,2016,2)) + 
  scale_color_manual(values=myPalette) +
  ggtitle("Death Rates by Age") +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())



```

